---
title: "Investigating the Formation of Co-authorship Ties in the Berlin-Bradenburg
  Political Science Network"
author: "Betty Lewis"
date: "2026-05-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
setwd("C://R_AST")
getwd()
load("berlin.RData")
dim(authorship)
coaut <- authorship %*% t(authorship)
dim(coaut)
seed <- 123
coautbin <- (coaut > 0) * 1
diag(coautbin) <- 0
library(igraph)
g <- graph_from_adjacency_matrix(coautbin, mode = "undirected", diag = FALSE)
personstwo <- persons[!duplicated(persons$name), ]
personstwo <- personstwo[1:227, ]
personstwo$genderbin <- ifelse(personstwo$gender == "female", 1, 0)
vertex_attr(g, "gender") <- personstwo$gender
vertex_attr(g, "genderbin") <- personstwo$genderbin
vertex_attr(g, "uni") <- personstwo$uni
```

INTRO 



To investigate the formation of co-authorship ties in the Berlin-Bradenburg political science network, first descriptive analysis is conducted. The network is very sparse, with an overall density of 0.00495, suggesting that many individuals are disconnected or only connected to few others. The network is made up of 127 components, which further indicates disconnect between groups in the social network. Many authors may prefer to work alone, or where collaboration does occur this is within a small select cluster of authors. Shared partner analysis illuminates this clustering. 33 dyads share one mutual partner, 12 share 2, and 3 share 2, with no more pairs shared beyond this. There are 21 triangles within the network, along with edgewise shared partner analysis, this suggests a tendency for clustering and potential for further triadic closure in the network. This is confirmed by the global transitivity score of 0.2079. Clustered co-authorship relationships can be partially explained by those with complementary skills wanting to share workloads, this leading to the formation of shared partners between those already collaborating (Lefield, P., 2018, p.515). These groups are often made up of collaborators in similar disciplines connected by shared research problems and interpretations of these. Crane terms these 'invisible colleges', smaller networks between individuals cross-institutionally who may meet at conferences and events, building acquaintanceship to increase the likelihood of collaboration (1972, in Leifeld, p. et al., 2017). Invisible colleges are a structural feature, formed through edgewise shared partners which explain the formation of this co-authorship network.


```{r global descriptive analysis, echo = FALSE, message = FALSE, warning = FALSE}
graph.density(g)
components(g)$no
library(ergm)
library(network)
net <- as.network(coautbin, directed = FALSE)
summary(net ~ esp(1:20))
summary(net ~ triangle)
detach("package:ergm", unload = TRUE)
detach("package:network", unload = TRUE)
transitivity(g, type = "global")
```

As well as invisible colleges, the presence of academic "stars" influence the collective research of a discipline, these being the most central researchers who connect clusters within the network (ibid). Moody (2004) posits that as well as shaping the progress of research disciplines, scientific stars 'attract a disproportionate level of research funding, high-profile appointments, and many students and collaborators' (p.214). Consequently, academic stars play an influential role in network structure. This is further investigated using degree centrality metrics. Global degree centrality is 1.1189, on average, individuals have only co-authored with one other individual. Below, a table of academic stars, those who have co-authored with a significantly above average number of individuals are displayed in a table. While 3 of these are men, only two are women. 



```{r homophily analysis, echo = FALSE, message = FALSE, warning = FALSE}
mean(degree(g))
degreevalues <- degree(g)
degreedf <- data.frame(name = personstwo$name, degree = degreevalues, gender = personstwo$gender)
top5degree <- degreedf[order(degreedf$degree, decreasing = TRUE), ][1:5, ]
rownames(top5degree) <- NULL
library(knitr)
kable(top5degree, 
      col.names = c("Author", "Degree", "Gender"),
      caption = "5 most frequent coauthors by degree centrality")
```

Cole and Zuckerman (1984) posit that academia presents a 'productivity puzzle' whereby men outperform women by publication and citation counts (p.218). This is more modernly discussed by Aksnes (et al., 2011) who found that female Norwegian academics were also likely to publish fewer publications and be cited less than their male counterparts. Bosco (et al., 2024) broadens this scope, positing that Southern European journal publications are dominated by men, with the routes taken by men and women in publishing diverging with every career stage (p.299). 



```{r}
assortativity(g,V(g)$genderbin)
#uni needs integer factors
uniint <- as.integer(factor(V(g)$uni))
assortativity_nominal(g, uniint, directed = FALSE)
```































*